\name{}: machine learning over interaction networks and ontology-based background knowledge predicts drug--target interactions

### 2021-04-27 paper/main.tex
.

While in this initial formulation of a GCNConv the node-wise update
step is defined by the sum over all neighboring node representations,
we can alter this formulation to other message passing schemes.  We
can rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$, and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We denote this
generalized layer type as \textsc{GENConv} following the notation of
PyTorch Geometric \cite{PytorchGeometric}.  While the reordering is
mainly important for numerical stability, this alteration also addresses
the vanishing gradient problem for deeper convolutional networks
\cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale and highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously. % , and the resulting information\todo{Which information?
  % Specify} hold no information on their own connectivity.
We further employ another mechanism to avoid redundancy and fading
signals in stacked graph convolutional networks, using residual
connections and a normalization scheme \cite{DeepGCN2019,
  DeeperGCN2020}.  The residual blocks are reusable and can be stacked
multiple times. % The structure of the GNN architecture we use is shown
% in Figure~\ref{fig:ResGraphConvBlocks}.

@@

\subsubsection{Combined prediction model}
Combining half-twin and graph convolutional neural networks, we map
all protein representations to their respective node features,
initializing the graph convolutional update steps. The resulting
representations are used for a similarity prediction.
% as presented in
% figure ~\ref{fig:HalfTwinNetwork}
When combining ontology and molecular features with or without the
graph model, we concatenate both protein features and both drugs
features, before plugging them into the graph model for the similarity
computation. An overview of the model architecture, combining both
feature types, is shown in Supplementary Figure
4. Here
the original representations are transformed by a dense layer and then
used as input of a stack (with height 3) of residual graph
convolutional blocks. The full eventual model of \name{} is shown in Supplementary Figure 4.

>>>>

While in this initial formulation of a GCNConv the node-wise update
step is defined by the sum over all neighboring node representations,
we can alter this formulation to other message passing schemes.  We
can rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$, and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We denote this
generalized layer type as \textsc{GENConv} following the notation of
PyTorch Geometric \cite{PytorchGeometric}.  While the reordering is
mainly important for numerical stability, this alteration also addresses
the vanishing gradient problem for deeper convolutional networks
\cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale and highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously. % , and the resulting information\todo{Which information?
  % Specify} hold no information on their own connectivity.
We further employ another mechanism to avoid redundancy and fading
signals in stacked graph convolutional networks, using residual
connections and a normalization scheme \cite{DeepGCN2019,
  DeeperGCN2020} as shown in Supplemtary Figure 3.  The residual
blocks are reusable and can be stacked multiple times.
% The structure of the GNN architecture we use is shown
% in Figure~\ref{fig:ResGraphConvBlocks}.

@@

\subsubsection{Combined prediction model}
Combining half-twin and graph convolutional neural networks, we map
all protein representations to their respective node features,
initializing the graph convolutional update steps. The resulting
representations are used for a similarity prediction.
% as presented in
% figure ~\ref{fig:HalfTwinNetwork}
When combining ontology and molecular features with or without the
graph model, we concatenate both protein features and both drugs
features, before plugging them into the graph model for the similarity
computation.  An overview of the model architecture, combining both
feature types, is shown in Supplementary Figure 4.  Here the original
representations are transformed by a dense layer and then used as
input of a stack (with height 3) of residual graph convolutional
blocks. %The full model of \name{} is shown in Supplementary Figure 4.
